---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mizerStarvation

<!-- badges: start -->
<!-- badges: end -->

This is an extension package for the mizer package
(https://sizespectrum.org/mizer/) to implement starvation mortality.

## Installation

<!--
You can install the released version of mizerStarvation from [CRAN](https://CRAN.R-project.org) with: 

``` r 
install.packages("mizerStarvation") 
```
-->

You can install the development version of mizerStarvation from GitHub
with
``` r
devtools::install_github("sizespectrum/mizerStarvation")
```

## Example

This is an artificial example just to illustrate usage. We start with the
North Sea model that comes with the mizer package.

```{r message=FALSE}
library(mizerStarvation)
library(tidyverse)
library(ggplot2)
params <- NS_params
plotSpectra(params, power = 2)
```

We add starvation mortality
```{r}
params <- setStarvation(NS_params, 10)
```

We decrease resource availability to create some starvation
```{r}
capacity <- getResourceCapacity(params)
capacity[w_full(params) > 0.1] <- 0
params <- setResource(params, resource_capacity = capacity)
initialNResource(params)[w_full(params) > 0.1] <- 0
```

We can calculate the starvation mortality for each species as a function of
size with `getStarvMort()`:
```{r}
starv_mort <- getStarvMort(params)
```
This returns a matrix. For plotting we turn this into a data frame with `melt()`
and send it to ggplot:
```{r}
ggplot(melt(starv_mort)) +
    geom_line(aes(x = w, y = value, colour = sp, linetype = sp), size = 1) +
    scale_x_log10() +
    xlab("Size [g]") +
    ylab("Starvation mortality [1/year]") +
    scale_colour_manual(values = params@linecolour) +
    scale_linetype_manual(values = params@linetype)
```

Of course now Saithe will go extinct, not only because of the starvation
mortality but also because it stops growing before maturity.
```{r}
sim <- project(params, t_max = 30)
plotBiomass(sim)
```

