---
title: "Extension Mechanism"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extension Mechanism}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(mizer)
library(mizerStarvation)
```

# Overview

The mizer package provides several routes for extending its behaviour without
editing the mizer source code. A full description of these mechanisms is given
in `vignette("extensions", package = "mizer")`. This vignette explains which
of those mechanisms mizerStarvation uses and why.

The starvation mortality added by this package is an *additional* mortality
source that runs alongside the standard background, fishing, and predation
mortalities. It is not a replacement for any of them. This calls for a
mechanism that *adds* to the mortality rate pipeline rather than replacing it.

# The `other_mort` mechanism

The `MizerParams` object carries a slot `other_mort`: a named list of rate
function names. During every time step, `getMort()` calls each function in
this list and adds its return value to the total per-capita mortality rate.
This is the correct tool when:

- the extra mortality source is dynamical (it responds to the current
  community state at each time step), and
- it should be superimposed on the standard mortality terms rather than
  replacing them.

mizerStarvation registers its rate function into this list:

```{r show-slot, eval = FALSE}
params@other_mort[["starvation"]]
#> [1] "starvMort"
```

The string `"starvMort"` is the name of the exported function
`mizerStarvation::starvMort()`. mizer looks up this name in the package
namespace at projection time, so the function must remain exported as long
as any saved model has starvation mortality enabled.

# Storing parameters in `species_params`

The only parameter that governs starvation mortality is `starv_coef`, one
value per species. This is stored as a column of the `species_params` data
frame:

```{r show-param, eval = FALSE}
params@species_params[["starv_coef"]]
```

Using `species_params` rather than `other_params()` is appropriate here
because `starv_coef` varies by species. It is also the standard place mizer
uses to keep per-species biological parameters, so it is automatically saved
and restored with the `MizerParams` object.

# Recording the extension

`setStarvation()` inserts an entry into the `params@extensions` named vector:

```{r show-extensions, eval = FALSE}
params@extensions[["mizerStarvation"]]
#> [1] "sizespectrum/mizerStarvation"
```

This records which extension package was used to set up the model. It makes
it possible for mizer to warn users when they try to use a saved model in a
session where mizerStarvation is not installed.

# How the pieces fit together

`setStarvation()` sets up all three pieces in one call:

```{r setStarvation, eval = FALSE}
params <- setStarvation(params, starv_coef = 10)
```

Passing `starv_coef = 0` reverses the change: it removes the `"starvation"`
entry from `other_mort`, removes the `starv_coef` column from
`species_params`, and removes `"mizerStarvation"` from `params@extensions`.

During a projection, at each time step and for each species and size class,
mizer calls `starvMort()`:

```{r starvMort-body, eval = FALSE}
starvMort <- function(params, n, n_pp, n_other, ...) {
    e <- getEReproAndGrowth(params, n = n, n_pp = n_pp, n_other)
    mu_s <- -t(t(e * params@species_params$starv_coef) / params@w)
    mu_s[e > 0] <- 0
    return(mu_s)
}
```

The function:

1. Calls the standard mizer function `getEReproAndGrowth()` to get the energy
   available for growth and reproduction, $E_r(w)$, at the current community
   state.
2. Computes the starvation mortality rate
   $$\mu_s(w) = \frac{-E_r(w)}{w} \times \texttt{starv\_coef}$$
   for size classes where the energy balance is negative ($E_r < 0$).
3. Sets mortality to zero for size classes with a positive energy balance.

The result is a species-by-size matrix with the same dimensions as
`initialN(params)`, which `getMort()` adds to the other mortality terms.

# Summary of extension points used

| Mechanism | Where | Purpose |
|---|---|---|
| `params@other_mort` | named list of rate function names | registers `starvMort` so `getMort()` calls it at each time step |
| `params@species_params$starv_coef` | column in species parameters | stores the per-species proportionality constant |
| `params@extensions` | named character vector | records that mizerStarvation set up this model |

This package does not use `setRateFunction()`, `setComponent()`,
`setExtMort()`, or S4 subclassing, because those mechanisms are either
designed for replacing rather than augmenting existing rates, or add more
infrastructure than this simple additive mortality term requires.
