---
title: "Extension Mechanism"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extension Mechanism}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(mizer)
library(mizerStarvation)
```

# Overview

The mizer package provides several routes for extending its behaviour without
editing the mizer source code. A full description of these mechanisms is given
in the [guide to extending
mizer](https://sizespectrum.org/mizer/articles/guide-extend-mizer.html), and
the [guide to creating an extension
package](https://sizespectrum.org/mizer/articles/guide-create-extension-package.html)
covers turning one into a package. This vignette explains which of those
mechanisms mizerStarvation uses and why.

The starvation mortality added by this package is an *additional* mortality
source that runs alongside the standard background, fishing, and predation
mortalities. It is not a replacement for any of them. This calls for a
mechanism that *adds* to the mortality rate pipeline rather than replacing it.

# The `other_mort` mechanism

A `MizerParams` object carries a register of extra mortality contributions:
a named list of rate function names, reached with `other_mort()`. During
every time step, `getMort()` calls each function in this list and adds its
return value to the total per-capita mortality rate. This is the correct
tool when:

- the extra mortality source is dynamical (it responds to the current
  community state at each time step),
- it should be superimposed on the standard mortality terms rather than
  replacing them, and
- it carries no state of its own that has to be updated from one time step
  to the next.

mizerStarvation registers its rate function into this list:

```{r show-slot, eval = FALSE}
other_mort(params)[["starvation"]] <- "starvMort"
other_mort(params)
#> $starvation
#> [1] "starvMort"
```

Assigning `NULL` to an entry removes it again, which is how
`setStarvation(params, starv_coef = 0)` switches starvation mortality off.

`other_encounter()` is the same mechanism for a contribution to the
encounter rate.

The string `"starvMort"` is the name of the exported function
`mizerStarvation::starvMort()`. mizer looks up this name in the package
namespace at projection time, so the function must remain exported as long
as any saved model has starvation mortality enabled.

# Storing parameters in `species_params`

The only parameter that governs starvation mortality is `starv_coef`, one
value per species. This is stored as a column of the `species_params` data
frame:

```{r show-param, eval = FALSE}
species_params(params)[["starv_coef"]]
```

Using `species_params` rather than `other_params()` is appropriate here
because `starv_coef` varies by species. It is also the standard place mizer
uses to keep per-species biological parameters, so it is automatically saved
and restored with the `MizerParams` object.

`setStarvation()` writes the column with the `given_species_params()`
replacement function rather than by assigning to the `species_params` slot
directly:

```{r set-param, eval = FALSE}
given_species_params(params)[["starv_coef"]] <- starv_coef
```

`starv_coef` is user input that mizer never calculates for itself, so
declaring it as a *given* species parameter records that provenance. It then
shows up in `given_species_params(params)` alongside the other parameters the
user supplied, rather than in `calculated_species_params(params)`.

# Registering with mizer on load

When mizerStarvation is loaded, its `.onLoad` hook registers the package with
mizer:

```r
.onLoad <- function(libname, pkgname) {
  mizer::registerExtension(pkgname, requirement = "sizespectrum/mizerStarvation")
}
```

This adds `mizerStarvation` to the session's extension chain. The `requirement`
string is a `pak` installation spec, so mizer can install the package
automatically if it is missing. The call is idempotent: reloading the package
(e.g. via `devtools::load_all()`) does not modify the chain a second time.

# Recording the extension in the params object

Registering the package in `.onLoad` only announces it to the *session*. A
`MizerParams` object additionally records which extensions were actually
applied to it, in its `extensions` slot:

```{r show-extensions, eval = FALSE}
params@extensions
#> $mizerStarvation
#>                    requirement                        version 
#> "sizespectrum/mizerStarvation"                        "0.2.0" 
```

`setStarvation()` writes that entry with `mizer::recordExtension()`:

```{r record-extension, eval = FALSE}
params <- recordExtension(params, "mizerStarvation", version = version)
```

`recordExtension()` takes the installation requirement from the session's
registered chain, preserves any entries other extensions have already made,
and inserts this one at the right position in the chain. This is deliberately
narrower than copying the whole of `getRegisteredExtensions()` into the
object: another extension package may well be loaded in the session without
having been applied to *this* model, and the slot is a record of what was
applied, not of what happened to be attached.

The `version` argument is the version of mizerStarvation whose object layout
the recorded component conforms to. `setStarvation()` stamps the installed
version only when it first adds starvation mortality to a model, and passes
`version = NULL` on subsequent calls so that an existing stamp is preserved.
Re-stamping on an ordinary modification would let an old object claim to be
current and so skip a migration it actually needs.

Storing this record lets mizer warn users who try to load a saved model in a
session where mizerStarvation is missing or out of date.

When starvation mortality is disabled by passing `starv_coef = 0`,
`setStarvation()` removes the `"mizerStarvation"` entry from
`params@extensions` again, so the saved object no longer requires the package.

# How the pieces fit together

`setStarvation()` sets up all three pieces in one call:

```{r setStarvation, eval = FALSE}
params <- setStarvation(params, starv_coef = 10)
```

Passing `starv_coef = 0` reverses the change: it removes the `"starvation"`
entry from `other_mort()`, removes the `starv_coef` column from
`species_params`, and removes `"mizerStarvation"` from `params@extensions`.

During a projection, at each time step and for each species and size class,
mizer calls `starvMort()`:

```{r starvMort-body, eval = FALSE}
starvMort <- function(params, n, n_pp, n_other, t = 0, ...) {
    e <- getEReproAndGrowth(params, n = n, n_pp = n_pp, n_other = n_other,
                            t = t)
    starv_coef <- species_params(params)[["starv_coef"]]
    mu_s <- -t(t(e * starv_coef) / params@w)
    mu_s[e > 0] <- 0
    mu_s
}
```

`getMort()` calls each registered function with `params`, `n`, `n_pp`,
`n_other`, `t` and the `component` name, so a rate function has to accept
`...` to absorb the arguments it does not use. Note that `t` is named
explicitly here and forwarded, so that starvation mortality is computed at the
same time as the rest of the mortality in models whose rates depend on time.

The function:

1. Calls the standard mizer function `getEReproAndGrowth()` to get the energy
   available for growth and reproduction, $E_r(w)$, at the current community
   state.
2. Computes the starvation mortality rate
   $$\mu_s(w) = \frac{-E_r(w)}{w} \times \texttt{starv\_coef}$$
   for size classes where the energy balance is negative ($E_r < 0$).
3. Sets mortality to zero for size classes with a positive energy balance.

The result is a species-by-size matrix with the same dimensions as
`initialN(params)`, which `getMort()` adds to the other mortality terms.

# Summary of extension points used

| Mechanism | Where | Purpose |
|---|---|---|
| `other_mort(params)` | named list of rate function names | registers `starvMort` so `getMort()` calls it at each time step |
| `species_params(params)$starv_coef` | column in species parameters, written with `given_species_params<-()` | stores the per-species proportionality constant |
| `params@extensions` | named list written with `recordExtension()` | records that mizerStarvation set up this model, and under which version |

This package uses none of the neighbouring mechanisms, and it is worth being
precise about why:

- `setRateFunction()` **replaces** mizer's calculation of a rate. Starvation
  mortality is an addition to the existing mortality, not a different way of
  computing it.
- `setExtMort()` sets a **fixed array**. Starvation mortality depends on the
  current energy balance, so it has to be recomputed from the community state
  at every time step.
- `setComponent()` is for a component with **dynamics of its own** — a
  detritus or carrion pool whose state is carried from one time step to the
  next. Starvation mortality keeps no state, so there would be nothing for
  `dynamics_fun` to update. A contribution registered with `setComponent()`
  belongs to its component: `getComponent()` reports it, `removeComponent()`
  removes it, and `other_mort()` does not list it.
- S4 subclassing changes what mizer's generic functions return. Nothing in
  mizer's output needs to change here, so this stays a metadata-only
  extension.
